A Formalization of Typed Lua

### 2015-06-14 doc/papers/paper/paper.tex
DLS paper: finished revision, "release candidate" for submission

\conferenceinfo{DLS '15}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

In this work, we give a formal description of Typed Lua, an
optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of record and object types via imperative update that does not sacrifice width subtyping, and combining flow typing with multiple assignment and with functions that return multiple values. While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

@@

The early error detection and better tools for structuring
programs are two advantages of statically typed languages that
lead programmers to migrate their code from a dynamically
typed to a statically typed language when their scripts
evolve into complex programs~\cite{tobin-hochstadt2006ims}.
This migration from dynamic to static typing usually involves
different languages that have distinct syntax and semantics,
requiring a complexe rewrite of existing programs instead of
incremental evolution.

@@

The first idioms that we focus on this paper is the use of assignment to build records and objects one field at a time, as in
the Lua fragment in Figure~\ref{fig:newpoint}, where the {\tt newpoint} function
is a constructor for objects with {\tt x} and {\tt y}
fields and a {\tt move} method.

@@

The {\tt newpoint} function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping.

@@

In order to find out how Lua is used in practice, to inform
the design of Typed Lua, we did a survey of existing Lua
libraries. Our corpus is the repository for the LuaRocks
package manager~\cite{hisham2013luarocks}, downloaded on
February 1st, 2014. We did not survey all of the Lua
scripts in the repository: scripts that were not compatible
with Lua 5.2 (the version of Lua at that time) were ignored,
and we also ignored machine-generated scripts and test scripts,
as those could skew the result towards non-idiomatic uses
of Lua. This left 2,598 scripts out of a total of 3,928, from
262 different projects. The scripts were parsed, and their
abstract syntax trees analysed to collect the data that we
show in the rest of this section.

@@

We also analysed expressions that access a table (130,448 instances), either to read a value out of it (86\% of such expressions) or to write a 
value into it (14\% of such expressions). Most of the expressions
that read a value out of a table use a literal key,
either a literal string (89\% of the reads) or a literal
number (4\% of the reads). Lua has syntactic sugar that
turns field accesses using ``dot'' notation into table
accesses through literal string keys.

@@

A Lua idiom that is specially problematic for static type
systems is overloading on the number and types and
multiple return values. Around 6\% of functions use
this idiom as a way to signal errors: in case of errors,
the function returns either {\tt nil} or {\tt false} plus
either an error message or error object instead of its usual
return values or throwing an exception. The caller then
tests the first returned value to check if an error occured
or not. Standard union types and flow typing cannot deal
with this, and Typed Lua introduces {\em projection types},
a way of tracking dependencies between types in a union of
tuples after a destructuring assignment.

@@

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{first-level types:}\\
& \;\; F & \textit{first-class types}\\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
F ::= & & \textsc{{\small first-class types:}}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \mathbf{void} & \textit{void type}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& | \; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\
%\multicolumn{3}{c}{}\\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\ 
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; T \times P & \textit{pair types}
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

Figure \ref{fig:typelang} presents the abstract syntax of
Typed Lua types. They are split into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear in 
assignments and function applications.

First-level types are
themselves split into {\em first-class types}, which can be
used in type annotations and most type constructors, and
{\em filter types} and {\em projection types}, special
types that only appear in support of flow typing.
Typed Lua includes projection types as a way to project
unions of tuple types into unions of first-level types,
and we detailt their use in Section~\ref{sec:projections} .

Types are ordered by a subtype relationship, so Lua values may belong to several distinct types. The details of the subtyping
relationship are given later in this section.

@@

The base types represent values that Lua tags as booleans,
strings, or numbers during runtime. Prior to version Lua 5.3 all Lua numbers were double-precision floating point numbers.
Lua 5.3 introduced the possiblity of having 64-bit integers
as numbers, with automatic promotion of integer values to
floating-point representation as needed. As automatic
promotion makes integers essentialy subtypes of floating-point numbers, Typed Lua uses $\Number$ as the type of floating-point numbers, and $\Integer$ as the type of integers.

The type {\bf void} is a type with no values, used by the
type system as a way to detect branches that are unreacheable
due to flow typing. This type is neutral with regards to
the type union operator: ${\bf void} \cup T = T \cup {\bf void} = T$ for any type $T$.

@@

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-class
types that end with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$. Lua pads sequences with {\tt nil} to
have as many values as needed, and in Typed Lua sequences
where the number of values is statically known have a $\Nil{*}$
tail to reflect this semantics.

For clarity, we use the symbol $\sqcup$ to represent the union between two different tuple types, and reserve $\cup$ for the
union of two first-class types.

@@

The consistent-subtyping relation is defined in terms
of subtyping, so we focus the discussion in this section
on characterising the subtyping relation.

@@

Figure~\ref{fig:subtuples} shows the subtyping rules for
second-level types. 
Tuple types are covariant, so the subtyping rule {\sc S-Pair} for pair types  is straightforward.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2} \;\;\;
	\senv \vdash V_{2} \subtype V_{1}}
{\senv \vdash V_{1} \subtype_{c} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{c} \Const \; V_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash \Const \; V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash V_{1} \subtype V_{2}}
{\senv \vdash V_{1} \subtype_{u} \Const \; V_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} V}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype V}
{\senv \vdash \Nil \subtype_{o} \Const \; V}
\end{array}
\end{array}
\]
\caption{Subtyping for fields}
\label{fig:subfield}
\end{figure}

@@

The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with a
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantess, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgement where this subtyping rule is used.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash T_{1} \subtype T_{2}}
{\senv \vdash \mu x_{1}.T_{1} \subtype \mu x_{2}.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash T_{1} \subtype [x \mapsto \mu x.T_{2}]T_{2}}
{\senv \vdash T_{1} \subtype \mu x.T_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.T_{1}]T_{1} \subtype T_{2}}
{\senv \vdash \mu x.T_{1} \subtype T_{2}}
\end{array}
\end{array}
\]
\caption{Subtyping for recursive types}
\label{fig:subrec}
\end{figure}

@@

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-class types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

@@

\begin{figure*}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
\overline{\env_1, \penv \vdash e_2: fix(V), \env_v}
	\;\;\;
	\overline{\env_v, \penv \vdash e_1: K, \env_f}
	\\
	T = \{\overline{K{:}V}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{K:V} \}_{unique}\\
	\env_{1}, \penv \vdash k:K^{new}, \env_{2} \;\;\;
	\overline{K^{new} \not\lesssim K} \;\;\;
	V^{new} = fix(V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[k] {<}V{>}:V^new,
	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdReadClose}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdReadFix}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:fix(F), \env[id \mapsto fix(F)]}
\end{array}
\end{array}
\]
\caption{Rules for Table Types}
\label{fig:tablerules}
\end{figure*}

@@

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgements for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.

@@

The $merge$ auxiliary function then combines output
environments, diverging in case of conflicting changes.
Fields can always be added to {\em unique} and {\em open}
tables in one enviroment but not the other. Two environments
can add a field with the same key, as long as they have
the same type or one is a supertype of the other, respecting
any restrictions with regards to depth subtyping that the table may have.

@@

Rule {\sc T-Id} is responsible for tracking
aliasing. Its auxiliary function $close$ is similar to
function $fix$ defined in Figure~\ref{fig:funcs},
except that it changes the tag of {\em unique} or
{\em open} tables to {\em closed}. Its purpose
is to construct a {\em closed} type for the alias,
if necessary. Its auxiliary function $open$ is similar
to $fix$ and $close$, except that it changes the type of
{\em unique} tables to {\em open}. Its purpose
is to track the fact that a $id$ now has an alias,
if necessary.

>>>>

\conferenceinfo{DLS '15}{October 25-30, 2015, Pittsburgh, PA, USA}
\copyrightyear{2015}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

In this work, we give a formal description of Typed Lua, an
optional type system for Lua, with a focus on two of its novel type system features: incremental evolution of imperative record and object types that is both lightweight and type-safe, and
the combination of flow typing, functions that return multiple values, and multiple assignment. While our type system is tailored to the features and idioms of Lua, its features can be adapted to other imperative scripting languages.
\end{abstract}

\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}[Type Systems]

\terms
Languages, Design

\keywords
Optional Type Systems, Gradual Typing, Lua

@@

The early error detection and better tools for structuring
programs are two advantages of statically typed languages that
lead programmers to migrate their code from a dynamically
typed to a statically typed language when their scripts
evolve into complex programs~\cite{tobin-hochstadt2006ims}.
This migration from dynamic to static typing usually involves
different languages that have distinct syntax and semantics,
requiring a complex rewrite of existing programs instead of
incremental evolution.

@@

The first idioms that we focus on this paper is the use of assignment to build records and objects one field at a time, as in
the Lua fragment in Figure~\ref{fig:newpoint}, where the {\tt new\_point} function
is a constructor for objects with {\tt x} and {\tt y}
fields and a {\tt move} method.

@@

The {\tt new\_point} function builds the object piece-by-piece,
from an initially empty table. Typed Lua
lets record and object types evolve incrementally 
via assignment while not sacrificing width subtyping.

@@

In order to find out how Lua is used in practice, to inform
the design of Typed Lua, we did a survey of existing Lua
libraries. Our corpus is the repository for the LuaRocks
package manager~\cite{hisham2013luarocks}, downloaded on
February 1st, 2014. We did not survey all of the Lua
scripts in the repository: scripts that were not compatible
with Lua 5.2 (the version of Lua at that time) were ignored,
and we also ignored machine-generated scripts and test scripts,
as those could skew the result towards non-idiomatic uses
of Lua. This left 2,598 scripts out of a total of 3,928, from
262 different projects. The scripts were parsed, and their
abstract syntax trees analyzed to collect the data that we
show in the rest of this section.

@@

We also analyzed expressions that access a table (130,448 instances), either to read a value out of it (86\% of such expressions) or to write a 
value into it (14\% of such expressions). Most of the expressions
that read a value out of a table use a literal key,
either a literal string (89\% of the reads) or a literal
number (4\% of the reads). Lua has syntactic sugar that
turns field accesses using ``dot'' notation into table
accesses through literal string keys.

@@

A Lua idiom that is specially problematic for static type
systems is overloading on the number and types and
multiple return values. Around 6\% of functions use
this idiom as a way to signal errors: in case of errors,
the function returns either {\tt nil} or {\tt false} plus
either an error message or error object instead of its usual
return values or throwing an exception. The caller then
tests the first returned value to check if an error occurred
or not. Standard union types and flow typing cannot deal
with this, and Typed Lua introduces {\em projection types},
a way of tracking dependencies between types in a union of
tuples after a destructuring assignment.

@@

\begin{figure*}[t]
$$
\begin{array}{rlr}
F ::= & & \textsc{first-level types:}\\
& \;\; L & \textit{literal types}\\
& | \; B & \textit{base types}\\
& | \; \Nil & \textit{nil type}\\
& | \; \Value & \textit{top type}\\
& | \; \Any & \textit{dynamic type}\\
& | \; \Self & \textit{self type}\\
& | \; F_{1} \cup F_{2} & \textit{union types}\\
& | \; S_{1} \rightarrow S_{2} & \textit{function types}\\
& | \; \{\overline{K:V}\}_{unique|open|fixed|closed} & \textit{table types}\\
& |\; x & \textit{type variables}\\
& | \; \mu x.F & \textit{recursive types}\\
%\multicolumn{3}{c}{}\\
L ::= & & \textsc{{\small literal types:}}\\
& \;\; \False \; | \; \True \; | \; {\it int} \; | \; {\it float} \; | \; {\it string} &\\
%\multicolumn{3}{c}{}\\
B ::= & & \textsc{{\small base types:}}\\
& \;\; \Boolean \; | \; \Integer \; | \; \Number \; | \; \String &\\\\
%\multicolumn{3}{c}{}\\
%\multicolumn{3}{c}{}\\
S ::= & & \textsc{second-level types:}\\
& \;\; P & \textit{tuple types}\\
& | \; S_{1} \sqcup S_{2} & \textit{unions of tuple types}\\
%\multicolumn{3}{c}{}\\
P ::= & & \textsc{{\small tuple types:}}\\
%& \;\; \Void & \textit{void type}\\
& | \; F{*} & \textit{variadic types}\\
& | \; F \times P & \textit{pair types}\\\\
\end{array}
$$
\caption{Type Language}
\label{fig:typelang}
\end{figure*}

\begin{figure*}[t]
$$
\begin{array}{rlr}
T ::= & & \textsc{{\small expression types:}}\\
& \;\; V & \\
& | \; \phi(F_{1},F_{2}) & \textit{filter types}\\
& | \; \pi_{i}^{x} & \textit{projection types}\\
E ::= & & \textsc{{\small expression list types:}}\\
& \;\; T{*} & \\
& | \; T \times E & \\
K ::= & & \textsc{{\small key types:}}\\
& \;\; L \; | \; B \; | \; \Value &\\
%\multicolumn{3}{c}{}\\
V ::= & & \textsc{{\small value types:}}\\
& \;\; F \; | \; \Const \; F &\\
R ::= & & \textsc{\small{filter result type:}}\\
& \;\; \mathbf{void} & \textit{void type}\\
& |\; F
\end{array}
$$
\caption{Special Types}
\label{fig:special}
\end{figure*}

Figure~\ref{fig:typelang} presents the abstract syntax of
the core Typed Lua types. They are split into two categories:
\emph{first-level types} and \emph{second-level types}.
First-level types represent first-class Lua values and
second-level types represent tuples of values that appear
as parameters and return values of functions.

Types are ordered by a subtype relationship, so Lua values may belong to several distinct types. The details of the subtyping
relationship are given later in Section~\ref{sec:subtyping}.

@@

The base types represent values that Lua tags as booleans,
strings, or numbers during runtime. Prior to version Lua 5.3 all Lua numbers were double-precision floating point numbers.
Lua 5.3 introduced the possibility of having 64-bit integers
as numbers, with automatic promotion of integer values to
floating-point representation as needed. As automatic
promotion makes integers essentially subtypes of floating-point numbers, Typed Lua uses $\Number$ as the type of floating-point numbers, and $\Integer$ as the type of integers.

The type {\bf void} is a type with no values, used by the
type system as a way to detect branches that are unreachable
due to flow typing. This type is neutral with regards to
the type union operator: ${\bf void} \cup T = T \cup {\bf void} = T$ for any type $T$.

@@

Second-level types are either tuple types or unions of tuple types, and represent sequences of values that appear as arguments, return values, and the left and right sides of
assignments. Tuple types are heterogeneous lists of first-level
types that end with a {\em variadic type}.
A variadic type $F{*}$ represents an unbounded sequence of values of type $F \cup \Nil$. Lua pads sequences with {\tt nil} to
have as many values as needed, and in Typed Lua sequences
where the number of values is statically known have a $\Nil{*}$
tail to reflect this semantics.

For clarity, we use the symbol $\sqcup$ to represent the union between two different tuple types, and reserve $\cup$ for the
union of two first-level types.

Figure~\ref{fig:special} presents some types that, while
they may not be present in type annotations, play some
special role in the type system. {\em Expression types}
denote the type of expressions in the type system, which,
in addition to first-level types, can also be
{\em filter types}, which are the result of flow typing,
and {\em projection types}, which are the result of
decomposing a union of tuples. {\em Expression list types}
denote the type of lists of expressions. {\em Key types}
and {\em value types} have also been mentioned above. Finally, {\em filter result types}
are also used in flow typing; the {\bf void} type is
a type with no values.

@@

The consistent-subtyping relation is defined in terms
of subtyping, so we focus the discussion in this section
on characterizing the subtyping relation.

@@

Figure~\ref{fig:subtuples} shows the subtyping rules for
second-level types. Expression list types have similar
subtyping rules. Tuple types are covariant, so the subtyping rule {\sc S-Pair} for pair types is straightforward.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-FieldClsMtoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2} \;\;\;
	\senv \vdash F_{2} \subtype F_{1}}
{\senv \vdash F_{1} \subtype_{c} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldClsItoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldClsMtoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{c} \Const \; F_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-FieldUnqMtoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{u} F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqItoM}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{u} F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldUnqItoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash \Const \; F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldUnqMtoI}\\
\dfrac{\senv \vdash F_{1} \subtype F_{2}}
{\senv \vdash F_{1} \subtype_{u} \Const \; F_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{S-FieldOpnNtoM}\\
\dfrac{\senv \vdash \Nil \subtype F}
{\senv \vdash \Nil \subtype_{o} F}
\end{array}
\;
\begin{array}{c}
\mylabel{S-FieldOpnNtoI}\\
\dfrac{\senv \vdash \Nil \subtype F}
{\senv \vdash \Nil \subtype_{o} \Const \; F}
\end{array}
\end{array}
\]
\caption{Subtyping for fields}
\label{fig:subfield}
\end{figure}

@@

The rationale for the permissiveness of rule {\sc S-TableUtoC} derives from the fact that reading a non-existing key just yields $\Nil$ instead of being an error in Lua semantics, plus the fact
that the type system guarantees that a table with a
{\em unique} type has no keys with a value that does not
belong to one of its key types, and does not have any
aliases. With these guarantees, a table with type $\{1{:}\String, 2{:}\String,
``x"{:}\Integer, ``y"{:}\Value\}_{unique}$ can be safely recast as a table with type 
$\{\Number{:}\String \cup \Nil, ``x"{:}\Number, ``z"{:}
\Number \cup \Nil\}_{closed}$ as long as it becomes
inaccessible with the original type, which Section~\ref{sec:tables} will show that is the case
for any type judgment where this subtyping rule is used.

@@

\begin{figure}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{S-Amber}\\
\dfrac{\senv[x_{1} \subtype x_{2}] \vdash F_{1} \subtype F_{2}}
{\senv \vdash \mu x_{1}.F_{1} \subtype \mu x_{2}.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-Assumption}\\
\dfrac{x_{1} \subtype x_{2} \in \senv}
{\senv \vdash x_{1} \subtype x_{2}}
\end{array}\\\\
\begin{array}{c}
\mylabel{S-UnfoldRight}\\
\dfrac{\senv \vdash F_{1} \subtype [x \mapsto \mu x.F_{2}]F_{2}}
{\senv \vdash F_{1} \subtype \mu x.F_{2}}
\end{array}
\;
\begin{array}{c}
\mylabel{S-UnfoldLeft}\\
\dfrac{\senv \vdash [x \mapsto \mu x.F_{1}]F_{1} \subtype F_{2}}
{\senv \vdash \mu x.F_{1} \subtype F_{2}}
\end{array}
\end{array}
\]
\caption{Subtyping for recursive types}
\label{fig:subrec}
\end{figure}

@@

The type $\Any$ is trivially consistent with all types, and all types
are consistent with $\Any$. A type is also consistent with
itself. Two union types $F_1 \cup F_2$ and $F_3 \cup F_4$ are
consistent if either $F_1 \sim F_3$ and $F_2 \sim F_4$ or
$F_1 \sim F_4$ and $F_2 \sim F_3$. Two function types
$S_1 \rightarrow S_2$ and $S_3 \rightarrow S_4$ are consistent
if and only if $S_1 \rightarrow S_3$ and $S_2 \rightarrow S_4$. Two table types are consistent if and only if they have the same key types, and the value types of each key type are consistent. Two recursive types are consistent if and only if their contents are consistent assuming that their type variables are consistent. Unions of tuple types follow the same rules as unions of first-level types. Variadic types are consistent if and only if their base types are consistent. 
Finally, pair types $F_1 \times P_1$ and $F_2 \times P_2$ are
consistent if and only if $F_1 \sim F_2$ and $P_1 \sim P_2$.

@@

\begin{figure*}[t]
\[
\begin{array}{c}
\begin{array}{c}
\mylabel{T-Constructor}\\
\dfrac{\begin{array}{c}
\overline{\env_1, \penv \vdash e_2: F, \env_v}
	\;\;\;
	\overline{\env_v, \penv \vdash e_1: K, \env_f}
	\\
	T = \{\overline{K{:}fix(F)}\}_{unique} \;\;\;
	\env_{2} = merge(\overline{\env_{f}}) \;\;\; wf(T)
	\end{array}}
{\env_{1}, \penv \vdash \{\;\overline{[e_{1}] = e_{2}}\;\}:T, \env_{2}}
\end{array}
\begin{array}{c}
\mylabel{T-NewFieldUnique}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{ \overline{K:V} \}_{unique}\\
	\env_{1}, \penv \vdash k:K^{new}, \env_{2} \;\;\;
	\overline{K^{new} \not\lesssim K} \;\;\;
	V^{new} = fix(V)
	\end{array}}
{\begin{array}{l}
	\env_{1}, \penv \vdash id[k] {<}V{>}:V^{new},
	\env_{2}[id \mapsto \{ \overline{K{:}V}, K^{new}{:}V^{new}\}_{unique}]
	\end{array}}
\end{array}\\\\
\begin{array}{c}
\mylabel{T-IdRead}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id:close(F), \env[id \mapsto open(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-IdWrite}\\
\dfrac{\env(id) = F}
{\env, \penv \vdash id_{l}:F, \env}
\end{array}
\begin{array}{c}
\mylabel{T-IndexId}\\
\dfrac{\begin{array}{c}
	\env_{1}(id) = \{\overline{K:V}\} \;\;\;
	\env_{1}, \penv \vdash e_{2}:K^e, \env_{2} \;\;\;
	 K^e \lesssim K_{i}
	\end{array}}
{\env_{1}, \penv \vdash id[e_{2}]:V_{i}, \env_{2}}
\end{array}
\\ \\
\begin{array}{c}
\mylabel{T-IndexExp}\\
\dfrac{\begin{array}{c}
	\env_{1}, \penv \vdash e_{1}:\{\overline{K:V}\}, \env_{2}\\
	\env_{2}, \penv \vdash e_{2}:K^e, \env_{3} \;\;\;
	 K^e \lesssim K_{i}
	\end{array}}
{\env_{1}, \penv \vdash e_{1}[e_{2}]:V_{i}, \env_{3}}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceClosed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,closed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto reopen(F)]}
\end{array}
\begin{array}{c}
\mylabel{T-CoerceFixed}\\
\dfrac{\env_{1}(id) \subtype F \;\;\; tag(F,fixed)}
{\env, \penv \vdash {<}F{>} \; id:F, \env[id \mapsto F]}
\end{array}\\
\begin{array}{c}
\mylabel{T-Function}\\
\dfrac{\begin{array}{c}
	closeall(\env_{1})[\overline{id \mapsto F}], \penv[\ret \mapsto S] \vdash s, \env_{2}\\
	\env_{3} = openset(\env_{1}, frv(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s)) \\
	\env_{4} = closeset(\env_{3}, fav(\mathbf{fun} \; (\overline{id{:}T}){:}S \; s))
	\end{array}}
{\env_{1}, \penv \vdash \mathbf{fun} \; (\overline{id{:}F}){:}S \; s:F_1 \times \ldots F_n \times \Void \rightarrow S, \env_{3}}
\end{array}
\end{array}
\]
\caption{Rules for Table Types}
\label{fig:tablerules}
\end{figure*}

@@

We present the typing rules as a deduction system for two typing relations, one for typing statements and another for typing expressions. Typing judgments for statements have the form $\env_{1}, \penv \vdash s, \env_{2}$, meaning that
given a type environment $\env_{1}$ and a {\em projection environment} $\penv$ we can prove that a statement $s$ produces a new type environment $\env_{2}$. Typing judgments for
expressions have the form $\env_{1}, \penv \vdash e : T, \env_{2}$, meaning that given a type environment $\env_{1}$
and a projection environment $\penv$ we can prove that an expression $e$ has type $T$ and produces a new type environment $\env_{2}$.

@@

The $merge$ auxiliary function then combines output
environments, diverging in case of conflicting changes.
Fields can always be added to {\em unique} and {\em open}
tables in one environment but not the other. Two environments
can add a field with the same key, as long as they have
the same type or one is a supertype of the other, respecting
any restrictions with regards to depth subtyping that the table may have.

@@

Rule {\sc T-IdRead} is responsible for tracking
aliasing. Its auxiliary function $close$ is similar to
function $fix$ defined in Figure~\ref{fig:funcs},
except that it changes the tag of {\em unique} or
{\em open} tables to {\em closed}. Its purpose
is to construct a {\em closed} type for the alias,
if necessary. Its auxiliary function $open$ is similar
to $fix$ and $close$, except that it changes the type of
{\em unique} tables to {\em open}. Its purpose
is to track the fact that a $id$ now has an alias,
if necessary.

Consider the following example, where we make an alias
of a {\em unique} table:
\begin{center}
\begin{tabular}{lll}
\multicolumn{3}{l}{$\mathbf{local} \; a:\{\}_{unique} = \{\} \; \mathbf{in}$}\\
& \multicolumn{2}{l}{$\mathbf{local} \; b:\{\}_{closed} = a \; \mathbf{in}a[``x"] \; {<}\String{>} = ``foo";$}
\end{tabular}
\end{center}

This example typechecks; after the initialization of $b$
the type of $a$ is {\em open} instead of {\em unique}, but
this still lets the assignment add a new field to $a$.
If we changed the tag of the type annotation of $b$ to
any other tag the example would not typecheck anymore,
as the type of expression $a$ is $\{\}_{closed}$.

If we did not change the tag on the type produced by $a$
we could break type safety, as we could declare $b$ with
tag {\em unique} or {\em open} and add fields that would
conflict with the type of the fields on $a$, but at
runtime both reference the same table.
